---
title: "Introduction to the PDM Rainfall-Runoff Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the PDM Rainfall-Runoff Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
library(reach.hydro)
```

The **Probability Distributed Model (PDM)** (Moore, 2007) is a conceptual
rainfall-runoff model used operationally for flood forecasting. It represents a
catchment as a probability distribution of soil moisture storage capacities, and
routes runoff through groundwater and surface linear reservoirs. This vignette
walks through a complete PDM workflow using synthetic forcing data.

---

## 1. Synthetic forcing data

We generate three years of daily rainfall and potential evapotranspiration (PET)
with a realistic seasonal cycle. Rainfall is highest in winter, PET highest in
summer.

```{r forcing}
set.seed(42)
n      <- 365 * 3
dates  <- seq.Date(as.Date("2020-10-01"), by = "day", length.out = n)
t_seq  <- seq_len(n)
season <- 0.5 + 0.5 * cos(2 * pi * (t_seq - 30) / 365)  # 1 in winter, 0 in summer

rain <- pmax(0, rnorm(n, mean = 2.5 * season, sd = 3))
pet  <- pmax(0, 3 * (1 - season) + rnorm(n, 0, 0.3))

cat(sprintf(
  "Forcing: %d days | mean rain = %.2f mm/day | mean PET = %.2f mm/day\n",
  n, mean(rain), mean(pet)
))
```

```{r plot-forcing, fig.cap = "Synthetic daily rainfall and PET over three years."}
par(mfrow = c(2, 1), mar = c(2, 4, 1, 1))

plot(dates, rain, type = "h", col = "steelblue", lwd = 0.6,
     ylab = "Rainfall (mm/day)", xlab = "")
plot(dates, pet,  type = "l", col = "darkorange", lwd = 1.2,
     ylab = "PET (mm/day)", xlab = "")

par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
```

---

## 2. Parameter setup

`pdm_params()` constructs and validates a `PdmParams` object. All Moore (2007)
Table 1 parameters have defaults; we only need to set the ones that differ from
defaults.

```{r params}
p <- pdm_params(
  dist  = "pareto",   # Appendix A — most commonly used distribution
  cmax  = 350,        # maximum soil storage capacity [mm]
  b     = 0.4,        # Pareto shape exponent
  St    = 20,         # soil tension storage threshold [mm]
  kg    = 150,        # recharge time constant [days]
  kb    = 200,        # baseflow time constant [days]
  m     = 1,          # groundwater exponent (1 = linear reservoir)
  k1    = 5,          # surface reservoir 1 time constant [days]
  k2    = 8           # surface reservoir 2 time constant [days]
)

print(p)
```

---

## 3. Running the model

`pdm()` returns a `ReachHydroResult` — a `data.table` subclass with 16 columns
covering all intermediate states and flow components.

```{r run}
res <- pdm(rain, pet, params = p)

print(res)
summary(res, warmup = 365)  # exclude first year from summary statistics
```

```{r plot-hydro, fig.cap = "Simulated flow separated into fast (Qf) and baseflow (Qb) components."}
par(mar = c(3, 4, 1, 1))
plot(dates, res$Q,  type = "l", col = "steelblue",  lwd = 1.2,
     ylab = "Flow (mm/day)", xlab = "")
lines(dates, res$Qf, col = "dodgerblue",  lwd = 0.8, lty = 2)
lines(dates, res$Qb, col = "sienna",      lwd = 0.8, lty = 2)
legend("topright",
       legend = c("Q total", "Qf fast", "Qb baseflow"),
       col    = c("steelblue", "dodgerblue", "sienna"),
       lty    = c(1, 2, 2), lwd = 1.2, bty = "n", cex = 0.85)
par(mar = c(5, 4, 4, 2) + 0.1)
```

```{r plot-stores, fig.cap = "Soil moisture (S) and groundwater store (Sg) over the simulation period."}
Smax <- attr(res, "Smax")

par(mfrow = c(2, 1), mar = c(2, 4, 1, 1))
plot(dates, res$S,  type = "l", col = "darkgreen", lwd = 1,
     ylim = c(0, Smax),
     ylab = "Soil moisture S (mm)", xlab = "")
abline(h = Smax, lty = 3, col = "grey50")
text(dates[10], Smax * 0.97, sprintf("Smax = %.0f mm", Smax),
     adj = 0, col = "grey50", cex = 0.8)

plot(dates, res$Sg, type = "l", col = "peru",      lwd = 1,
     ylab = "Groundwater Sg (mm)", xlab = "")
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
```

---

## 4. Goodness-of-fit against a reference run

To demonstrate performance metrics we create a reference ("observed") flow series
by running the model with slightly different parameters and adding a small
noise term.

```{r gof}
p_ref  <- pdm_params(dist = "pareto", cmax = 320, b = 0.35, St = 15,
                     kg = 130, kb = 180, m = 1, k1 = 5, k2 = 8)
set.seed(7)
obs_q  <- pdm(rain, pet, params = p_ref)$Q +
          pmax(0, rnorm(n, 0, 0.005))

# Exclude the first year as warm-up
metrics <- gof_metrics(obs_q, res$Q, warmup = 365)
print(round(metrics, 3))
```

The NSE and KGE are moderately high because the two parameter sets are
close but not identical — realistic for an uncalibrated model.

---

## 5. Comparing capacity distributions

`compare_distributions()` runs the PDM with all five supported capacity
distributions and returns a summary table. We run it on the first year only
to keep the example quick.

```{r compare-dists}
tbl <- compare_distributions(
  rain[1:365], pet[1:365],
  obs_q   = obs_q[1:365],
  warmup  = 0L
)
print(tbl[, .(distribution, Smax_mm, mean_Q, BFI, NSE, KGE)])
```

---

## 6. Calibration

`calibrate_pdm()` uses Nelder-Mead optimisation to find the parameter set that
maximises a goodness-of-fit metric against observed flow. We calibrate on the
first two years and evaluate on the third.

> **Note:** `maxit = 300` is used here for vignette build speed. Production
> calibration runs should use the default `maxit = 3000`.

```{r calibrate}
n_cal <- 365 * 2

cal <- calibrate_pdm(
  rain    = rain[1:n_cal],
  pet     = pet[1:n_cal],
  obs_q   = obs_q[1:n_cal],
  dist    = "pareto",
  warmup  = 365L,
  maxit   = 300L     # increase to 3000 for production
)

cat(sprintf("Calibration  NSE = %.3f  KGE = %.3f  PBIAS = %.1f%%\n",
            cal$nse, cal$kge, cal$pbias))
print(cal$params)
```

Evaluate on the held-out third year:

```{r eval-holdout}
idx_val <- seq(n_cal + 1L, n)
res_cal <- pdm(rain[idx_val], pet[idx_val], params = cal$params)
val_m   <- gof_metrics(obs_q[idx_val], res_cal$Q)
cat(sprintf("Validation   NSE = %.3f  KGE = %.3f  PBIAS = %.1f%%\n",
            val_m["nse"], val_m["kge"], val_m["pbias"]))
```

---

## 7. State updating for operational forecasting

In operational use the PDM is initialised from recent observations rather than
from a cold start, removing the need for a multi-year burn-in. `update_pdm_states()`
runs the model over a recent assimilation window and adjusts the routing stores
to match the observed flow at the forecast origin.

```{r state-update}
# Assimilation window: last 30 days before the forecast origin
win       <- 30L
win_idx   <- seq(n - win, n - 1L)   # days 1066-1095 (end of year 3)
fcast_idx <- seq(n, min(n + 13L, n))  # use last day as "origin" for demo

upd <- update_pdm_states(
  rain   = rain[win_idx],
  pet    = pet[win_idx],
  obs_q  = obs_q[win_idx],
  params = cal$params,
  method = "inversion"
)

print(upd)
```

Run a 14-day forecast from the updated states:

```{r forecast, fig.cap = "14-day forecast (blue) against reference flow (orange) starting from updated states."}
set.seed(55)
n_fcast    <- 14L
rain_fcast <- pmax(0, rnorm(n_fcast, mean = 1.5, sd = 2))
pet_fcast  <- pmax(0, rep(1.8, n_fcast) + rnorm(n_fcast, 0, 0.15))
obs_fcast  <- pdm(rain_fcast, pet_fcast, params = p_ref)$Q

fcast <- pdm(rain_fcast, pet_fcast, params = cal$params,
             S0   = upd$S0,
             Sg0  = upd$Sg0,
             Ss10 = upd$Ss10,
             Ss20 = upd$Ss20)

fcast_dates <- seq.Date(tail(dates, 1) + 1L, by = "day", length.out = n_fcast)

par(mar = c(3, 4, 2, 1))
plot(fcast_dates, obs_fcast, type = "l", col = "darkorange", lwd = 1.5,
     ylim = range(c(obs_fcast, fcast$Q)) * c(0.8, 1.2),
     ylab = "Flow (mm/day)", xlab = "", main = "14-day forecast")
lines(fcast_dates, fcast$Q, col = "steelblue", lwd = 1.5, lty = 2)
legend("topright",
       legend = c("Reference", "Forecast"),
       col    = c("darkorange", "steelblue"),
       lty    = c(1, 2), lwd = 1.5, bty = "n", cex = 0.9)
par(mar = c(5, 4, 4, 2) + 0.1)
```

---

## References

Moore, R.J. (2007). The PDM rainfall-runoff model.
*Hydrology and Earth System Sciences*, **11**, 483–499.
<https://doi.org/10.5194/hess-11-483-2007>
